Value-Suppressing Uncertainty Maps

### 2017-03-31 paper/template.tex
Inline formulas

\begin{enumerate}
	\item \textbf{Select a visual variable to encode value}. In this paper, this variable is position along some established color scale.
	\item \textbf{Select a visual variable to encode uncertainty}. We prefer variables that have a somewhat high semiotic connection with uncertainty (such as lightness and size), but still have a reasonable number of perceptually discriminable layers (as opposed to other, more intuitive channels such as blur or sketchiness). In most cases in this paper, we encode uncertainty as interpolation to white or as glyph size.
	\item \textbf{Select a discriminability threshold}. Choose some minimum distance, in a perceptual space, between marks. For many visual variables (such as length or hue), the perceptual discriminability between marks has been assessed in prior literature, and so the only step is to choose a value in terms of the just noticeable distance (JND). For less common visual variables, a perceptual model may be built empirically (as in Demiralp \ea~\cite{demiralp2014learning}).
	\item \textbf{Choose how quickly categories \emph{degrade}}. That is, if there are $n$ bins at some uncertainty level $m$, how many bins should there be when uncertainty increases to level $m+1$? In \figref{fig:example}, this function is $\left \lfloor {\frac{n}{2}}\right \rfloor$, meaning that the number of value bins decreases by half for each level of uncertainty bins. If there is too much degradation, then there are fewer levels of uncertainty. If there is no degradation, the resulting map is just a traditional bivariate map.
	\item \textbf{Choose the largest number of initial categories $n$ such that the discriminability threshold is met.} For each pair of marks in the map, perform a pairwise comparison of discriminability. If no marks are perceptually ambiguous by the standards of your threshold, the map is complete. Otherwise, decrement $n$.
\end{enumerate}

@@

Using a public data set from the U.S. Bureau of Transportation Statistics~\cite{bts}, we created a heatmap showing average flight delay for different times of the day and days of the week for U.S. carriers in January 2017. We use standard error ($\frac{\sigma}{\sqrt{n}}$) as our measure of uncertainty. Starting from an minimum discriminability threshold of $18$ units in CIELAB space, we created two alternate heatmaps of this delay information. \figref{fig:airline2d} shows a traditional bivariate map, whereas \figref{fig:airlineVsum} shows a VSUM generated under the same constraints.

@@

For each task, participants saw one of three types of heatmaps. Either two \textbf{juxtaposed} maps side-by-side, with separate encodings for value and uncertainty, a traditional \textbf{2D bivariate} map, where value and uncertainty are orthogonally encoded, or a \textbf{VSUM}. \figref{fig:conditions} shows examples of these factor levels. Other factors included the size of the map (either a 4x4 or 8x8 grid), and (for the prediction task), the question framing (attack or defense). All factors were presented within subjects. Secondary factors were evenly blocked across stimuli. Participants reported fatigue effects in piloting, and so we limited the number of replicates in this study: participants saw 18 stimuli (6 for each map type) for the identification task, and 12 (4 for each map type) for the prediction task.

>>>>

\begin{enumerate}
	\item \textbf{Select a visual variable to encode value}. In this paper, this variable is position along some established color scale.
	\item \textbf{Select a visual variable to encode uncertainty}. We prefer variables that have a somewhat high semiotic connection with uncertainty (such as lightness and size), but still have a reasonable number of perceptually discriminable layers (as opposed to other, more intuitive channels such as blur or sketchiness). In most cases in this paper, we encode uncertainty as interpolation to white or as glyph size.
	\item \textbf{Select a discriminability threshold}. Choose some minimum distance, in a perceptual space, between marks. For many visual variables (such as length or hue), the perceptual discriminability between marks has been assessed in prior literature, and so the only step is to choose a value in terms of the just noticeable distance (JND). For less common visual variables, a perceptual model may be built empirically (as in Demiralp \ea~\cite{demiralp2014learning}).
	\item \textbf{Choose how quickly categories \emph{degrade}}. That is, if there are $n$ bins at some uncertainty level $m$, how many bins should there be when uncertainty increases to level $m+1$? In \figref{fig:example}, this function is $\left \lfloor n / 2 \right \rfloor$, meaning that the number of value bins decreases by half for each level of uncertainty bins. If there is too much degradation, then there are fewer levels of uncertainty. If there is no degradation, the resulting map is just a traditional bivariate map.
	\item \textbf{Choose the largest number of initial categories $n$ such that the discriminability threshold is met.} For each pair of marks in the map, perform a pairwise comparison of discriminability. If no marks are perceptually ambiguous by the standards of your threshold, the map is complete. Otherwise, decrement $n$.
\end{enumerate}

@@

Using a public data set from the U.S. Bureau of Transportation Statistics~\cite{bts}, we created a heatmap showing average flight delay for different times of the day and days of the week for U.S. carriers in January 2017. We use standard error $\left(\sigma / \sqrt{n}\right)$ as our measure of uncertainty. Starting from an minimum discriminability threshold of $18$ units in CIELAB space, we created two alternate heatmaps of this delay information. \figref{fig:airline2d} shows a traditional bivariate map, whereas \figref{fig:airlineVsum} shows a VSUM generated under the same constraints.

@@

For each task, participants saw one of three types of heatmaps. Either two \textbf{juxtaposed} maps side-by-side, with separate encodings for value and uncertainty, a traditional \textbf{2D bivariate} map, where value and uncertainty are orthogonally encoded, or a \textbf{VSUM}. \figref{fig:conditions} shows examples of these factor levels. Other factors included the size of the map (either a 4x4 or 8x8 grid), and (for the prediction task), the question framing (attack or defense). Secondary factors were evenly blocked across stimuli, resulting in a 3x2 within subjects design for the identification task, and a 3x2x2 within subjects design for the prediction task. Participants reported fatigue effects in piloting, and so we limited the number of replicates in this study: participants saw 18 stimuli (6 for each map type) for the identification task, and 12 (4 for each map type) for the prediction task.
